Topic categorization in Portuguese news articles\\ \thanks{Fundação para a Ciência e Tecnologia}

### 2018-01-03 article.tex
Methods

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{listings}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\section{Introduction}
The last decades have witnessed a fast increase on the rate of publication
of digital text documents. Traditional document types, such as news
articles, scientific papers or books are now published online together
with new formats, such as blog posts or tweets, each having thousands
or millions of new documents published each day.

And it was not only the publication step which has moved to the
digital world; in fact, most often nowadays the whole document
lifecycle happens digitally, with virtual tools available for
preparing, writing, styling, publishing and sharing.

Having the entire workflow happening within the digital world presents
some opportunities when compared to the traditional process. In
particular, due to todays processing power, tasks related to the
manipulation of the information contained within these documents
(searching, compiling, annotating, sharing, \dots) can now be
performed automatically and targetting a large amount of articles.

@@

Online news articles first appeared as reprints from traditional
newspapers; nowadays, however, they represent now the primary source
of news for some segments of the population, both in developed and
developing countries (whether consumed directly in the newspaper
website, or indirectly (e.g. through a social media application or a
feed catcher).

Unofficially know as \emph{the fourth branch of government}, the press
plays a vital role within our society, keeping us informed regarding
the current state of affairs (at a local and global scale) and acting
as a watchdog for the other three powers (legislative, executive and
judicial). The (lack of) freedom of press and access to the news in a
given country is even often considered an indicator of a lack of
democracy\cite{goode2009social}.

As such, improving the ways citizens can access the information (view
it, query it and search it) contained in news articles has the
potential to contribute for a more informed society and, ultimately, a
better society.

@@

@@

Thus, an automated way of categorizing news articles could solve some
of these problems and decrease the burden of this task.

@@

In this article, we describe the preliminary results obtained in
developing a classifier to categorize news articles, and in developing
a similar classifier with restrictions added to the maximum size of
the lexicon used % TODO

@@

\subsection{The dataset}
We gathered a dataset of news articles published in
\textit{Observador}\footnote{\url{http://observador.pt}}, one of the
main Portuguese newspapers, and which stands out from the others for
being fairly young (it was created in May 2014) and for existing
exclusively online. The initial dataset comprised 42.475 entries, from
which we used only a subset, for reasons later described.

We gathered all the categories used by Observador, and ordered them
from the most common to the least common. We selected the ones which
had more than 1.000 articles in our dataset, and reduced our original
dataset to include only articles from these categories.
Figure~\ref{fig:articles} presents an overview of the selected
categories and the number of articles available for each one.

@@

\begin{lstlisting}[caption={Example of JSON representation of an article},label={lst:article}, extendedchars=true]
{
  Type: "sapo.obj.creativework.article",
  Source: {
      Name : "Observador"
  },
  Pretitle: "Benfica",
  Title: "Ruben Amorim com rotura total do ligamento cruzado",
  Author: {
      Name: "Observador"
  },
  Tags: [
      "benfica",
      "desporto",
      "futebol",
      "ruben amorim"
  ],
  PublishDate: ISODate("2014-08-25T18:33:00Z"),
  Lead: "Depois de Fejsa, mais uma baixa. O internacional português [...]",
  Body: "<p>O pior cenário confirmou-se. O Benfica informou esta segunda-feira [...]",
  URL: "http://observador.pt/2014/08/25/ruben-amorim-com-rotura-total-ligamento-cruzado/",
  CategoryPaths: [
      "Desporto"
  ],
  Domain: "observador.pt",
  Language: "pt_PT",
  ...
}
\end{lstlisting}

\subsection{Preprocessing the articles}

Originally, the dataset was obtained as large (more than 2.5 million
entries) MongoDB collection, containing articles from several
Portuguese and international newspapers. The process required to
transform this collection into data our classifiers could process
required: querying the database, exporting the news articles and
splitting them into a train and an evaluation datasets.

The database query selected articles from \texttt{Observador} where
the body had a length greater than 100 characters (to
discard some malformed articles which had an empty body or a body
composed of only a few words), and the categories included at least
one of the categories we previously selected.

@@

The files were then stored in folders, separated by the category.
Furthermore, for each category, 700 articles were allocated to the
train set, and 200 to the evaluation set.

These preprocessing tasks were accomplished using Bash and Node.js
scripts.

\subsection{Classification algorithms}

\subsubsection{Decision tree}
\subsubsection{k-nearest neighbors}
\subsubsection{Naive Bayes}
\subsubsection{Neural network}
\subsubsection{Support vector machine}

\subsection{Most informative terms}
% for real-time topic suggestion (for an IDE for example)
% memory limit
% 100 terms? 1000 terms?

@@

\begin{table}[htbp]
    \caption{Categories confusion matrix (k-nearest neighbors)}
\begin{center}
\begin{tabular}{l|rrrrrrrr}
              & ciencia     & cultura     & desporto     & economia     & mundo       & musica       & pais        & politica \\\hline
ciencia       & \textbf{64} & 18          & 25           & 29           & 21          & 37           & 4           & 2 \\
cultura       & 4           & \textbf{37} & 16           & 9            & 19          & 107          & 3           & 5 \\
desporto      & 1           & 4           & \textbf{130} & 8            & 19          & 35           & 2           & 1 \\
economia      & 0           & 6           & 9            & \textbf{127} & 12          & 32           & 1           & 13 \\
mundo         & 9           & 8           & 25           & 31           & \textbf{82} & 32           & 1           & 12 \\
musica        & 0           & 6           & 1            & 1            & 0           & \textbf{180} & 0           & 1 \\
pais          & 8           & 15          & 24           & 42           & 17          & 64           & \textbf{14} & 16 \\
politica      & 1           & 8           & 7            & 36           & 19          & 24           & 3           & \textbf{102} \\

\end{tabular}
\label{tab1}
\end{center}
\end{table}
% >\\  error.rate.knn
% [1] 0.5368156
\subsection{Evaluation}
\subsection{Limited lexicon}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{float}
\usepackage{textcomp}
\usepackage{listings}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage[utf8x]{inputenc}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\section{Introduction}
Online news articles first appeared as reprints from traditional
newspapers; nowadays, however, they represent now the primary source
of news for some segments of the population, both in developed and
developing countries (whether consumed directly in the newspaper
website, or indirectly e.g. through a social media application or a
feed catcher).

Unofficially know as \emph{the fourth branch of government}, the press
plays a vital role within our society, keeping us informed about
the current state of affairs (at a local and global scale) and acting
as a watchdog for the other three branches (legislative, executive and
judicial). The (lack of) freedom of press and access to the news in a
given country is even often considered an indicator of a lack of
democracy\cite{goode2009social}.

As such, improving the ways citizens can access the information (view
it, query it and search it) contained in news articles has the
potential to contribute for a more informed and, ultimately, a
better society.

On the other hand, the last decades have witnessed a fast increase on
the rate of publication of digital text documents. Traditional
document types, such as news articles, scientific papers or books are
now published online together with new formats, such as blog posts or
tweets, each having thousands or millions of new documents published
each day.

And it was not only the publication step which has moved to the
digital world; in fact, most often nowadays the whole document
lifecycle happens digitally, with virtual tools available for
researching, writing, styling, publishing and sharing.

Having the entire workflow happening within the digital world presents
some opportunities when compared to the traditional process. In
particular, due to the current processing power commonly available,
tasks related to the manipulation of the information contained within
these documents (searching, compiling, annotating, sharing, \dots) can
now be performed automatically and targetting a large amount of
articles.

@@

@@

\begin{figure}[htbp]
\centerline{\includegraphics[width=.97\columnwidth]{imgs/wireframe}}
\caption{Category classification and suggestion on the client side}
\label{fig:wireframe}
\end{figure}

@@

Thus, an automated way of categorizing news articles could solve some
of these problems and decrease the burden of this task. Additionally,
a lightweight version of such a classifier could be implemented on the
client side code of a web application, for example, allowing the
categorization to happen in real time (i.e. as the author types in the
article text). Figure~\ref{fig:wireframe} presents a suggestion of
how this feature could look like if implemented on a web application.

@@

In this article, we describe the preliminary results obtained in
developing a classifier to categorize news articles using a previously
manually categorized corpus. Additionally, we evaluate the possibility
of implementing such a classifier as lightweight as possible to allow
it to run on the client side of a web application.

@@

\subsection{The dataset}
We gathered a dataset of news articles published in
\textit{Observador}\footnote{\url{http://observador.pt}}, one of the
main Portuguese newspapers, which stands out from the others for being
fairly young (it was created in May 2014) and for existing exclusively
online. The initial dataset comprised 42.475 entries, the most recent
ones dated from November 2016, from which we used only a subset, for
reasons later described.

We gathered all the categories used by Observador, and ordered them
from the most common to the least common. We selected the ones which
had more than 1000 articles in our dataset, and reduced our original
dataset to include articles from these categories only.
Figure~\ref{fig:articles} presents an overview of the selected
categories and the number of articles available for each one.

\begin{lstlisting}[float,floatplacement=H,caption={Example of JSON representation of an article},label={lst:article}, extendedchars=true]
{
  Type: "sapo.obj.creativework.article",
  Source: {
      Name : "Observador"
  },
  Pretitle: "Benfica",
  Title: "Ruben Amorim com rotura total do ligamento cruzado",
  Author: {
      Name: "Observador"
  },
  Tags: [
      "benfica",
      "desporto",
      "futebol",
      "ruben amorim"
  ],
  PublishDate: ISODate("2014-08-25T18:33:00Z"),
  Lead: "Depois de Fejsa, mais uma baixa. O internacional português [...]",
  Body: "<p>O pior cenário confirmou-se. O Benfica informou esta segunda-feira [...]",
  URL: "http://observador.pt/2014/08/25/ruben-amorim-com-rotura-total-ligamento-cruzado/",
  CategoryPaths: ["Desporto"],
  Domain: "observador.pt",
  Language: "pt_PT",
  ...
}
\end{lstlisting}

@@

\subsection{Preprocessing the articles}
\label{sec:preproc}

Originally, the dataset was obtained as large MongoDB collection (more
than 2.5 million entries), containing articles from several Portuguese
and international newspapers. The process needed to transform this
collection into data our classifiers could process required querying
the database, exporting the news articles and splitting them into a
train and an evaluation datasets.

The database query selected articles from \texttt{Observador} where
the body had a length greater than 100 characters (to
discard some malformed articles which had an empty body or a body
composed of only a few words), and the categories included at least
one of the most common categories.

@@

The files were then stored in folders, separated by the category.
Furthermore, for each category, 700 articles were allocated to the
train set, and 200 to the evaluation set. These preprocessing tasks
were accomplished using Bash and Node.js scripts.

Both datasets were then loaded into R using the
\texttt{tm}\footnote{\url{https://cran.r-project.org/web/packages/tm/}}
package. Each was then passed to a function responsible for
preprocessing the text of the articles:

\begin{itemize}
    \item the text was converted to lowercase characters
    \item the Portuguese stopwords were removed
    \item diacritics were converted to their normalized form (e.g.
        \framebox{à} $\rightarrow$ \framebox{a})
    \item punctuation signs were removed
    \item numbers were removed
    \item word were stemmed (e.g. \framebox{conseguiram} $\rightarrow$
        \framebox{consegu})
    \item whitespace was removed and the text was tokenized
\end{itemize}

%TODO
A document-term matrix was then calculated using the
\texttt{DocumentTermMatrix} method from the \texttt{tm} package. 
For each document in the train set the matrix included all the
terms which met the following requirements:

\begin{itemize}
    \item the term length was between 3 and 30 characters (to discard
        things like URLs or badly tokenized sentences)
    \item the term appeared in the document at least twice
    \item the term appeared at least in 10 documents
\end{itemize}
The algorithm used to weigh the terms in the matrix was the
\textit{tf-idf}, and resulted in 3234 terms.

%TODO matrix stats(sparcity etc)
% inspect(dtm)

The final step was to remove the sparse terms from the matrix.
This allows to discard terms which might be too specific of the train
set and which might negatively influence the performance of the
classifiers by overfitting them to the train set. Additionally, a
smaller list of terms reduces the execution time of both training and
applying the classifier, and the memory footprint of the classifier.
However, while discarding sparse terms we might end up removing
relevant terms and increase our classifiers error rate.

We used the \texttt{removeSparseTerms} function from the \texttt{tm}
package, and produced three distinct lists of terms:
\begin{description}
    \item[$TS_{99}$] contained the terms with a sparcity level below 99\% (912 terms)
    \item[$TS_{95}$] contained terms whose sparcity was under 95\% (139 terms)
    \item[$TS_{90}$] contained terms with 90\% or less sparcity (45 terms)
\end{description}

\subsection{Classification}

To create the classifiers, we used 5 well knows algorithms: decision
tree (DT), k-nearest neighbors (KNN), naive Bayes (NB), neural network
(NN) and support vector
machine -- with radial kernel (SVM\textsubscript{RK}) and linear
kernel (SVM\textsubscript{LK}).

We trained each of the classifiers three different times, one for each
list of terms ($TS_{99}$, $TS_{95}$ and $TS_{90}$). Then we evaluated
each trained classifier using the test dataset.

The test dataset was preprocessed in a way similar to the train set
(described in the previous section,~\ref{sec:preproc}),
but the final list of terms in each test document was restricted to the
terms considered for the train set.

@@

\begin{table}[htbp]
    \caption{Error rates obtained for each algorithm and each list of
    terms}
\begin{center}
\begin{tabular}{l|rrrrrr}
                 & DT            & KNN           & NB   & NN   & SVM\textsubscript{RK} & SVM\textsubscript{LK} \\ \hline
    $TS_{90}$    & \textbf{0.60} & 0.71          & 0.85 & 0.61 & 0.88                  & 0.88                  \\
    $TS_{95}$    & 0.70          & \textbf{0.56} & 0.83 & 0.58 & 0.87                  & 0.87                  \\
    $TS_{99}$    & 0.70          & \textbf{0.35} & 0.87 & 0.76 & 0.87                  & 0.87                  \\
\end{tabular}
\label{error}
\end{center}
\end{table}
\begin{table}[htbp]
    \caption{Categories confusion matrix (k-nearest neighbors)}
\begin{center}
\begin{tabular}{l|rrrrrrrr}
& \rotatebox{75}{ciencia} & \rotatebox{75}{cultura} &
    \rotatebox{75}{desporto} & \rotatebox{75}{economia} &
    \rotatebox{75}{mundo} & \rotatebox{75}{musica} &
    \rotatebox{75}{pais} & \rotatebox{75}{politica} \\\hline
ciencia       & \textbf{64} & 18          & 25           & 29           & 21          & 37           & 4           & 2 \\
cultura       & 4           & \textbf{37} & 16           & 9            & 19          & 107          & 3           & 5 \\
desporto      & 1           & 4           & \textbf{130} & 8            & 19          & 35           & 2           & 1 \\
economia      & 0           & 6           & 9            & \textbf{127} & 12          & 32           & 1           & 13 \\
mundo         & 9           & 8           & 25           & 31           & \textbf{82} & 32           & 1           & 12 \\
musica        & 0           & 6           & 1            & 1            & 0           & \textbf{180} & 0           & 1 \\
pais          & 8           & 15          & 24           & 42           & 17          & 64           & \textbf{14} & 16 \\
politica      & 1           & 8           & 7            & 36           & 19          & 24           & 3           & \textbf{102} \\

\end{tabular}
\label{confmat}
\end{center}
\end{table}
